Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-03-22 ms/S2_appendix.tex
minor tweaks to ms

\renewcommand\thefigure{S\arabic{figure}}

@@

This section presents alternate versions of
the figures from the main text showing time dynamics and summary
statistics in terms of expected progression time to AIDS and per-year transmission probability
rather than \Lspvl\.

>>>>

\renewcommand\thefigure{\Alph{figure}}

@@

This section presents alternate versions of
the figures from the main text showing time dynamics and summary
statistics in terms of expected progression time to AIDS and per-year transmission probability
rather than \Lspvl.

### 2017-03-22 ms/ms.tex
minor tweaks to ms

In models that allow extra-pair contact,
the uncoupled and extra-couple transmission rates (i.e., the rates of
transmission among people outside of a stable partnership, or between
people inside of a stable partnership and people other than their
partner) are scaled by
multiplying the within-couple transmission rate $\beta$ by the contact
ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1).

@@

Lastly, we incorporated heterogeneity in sexual activity into the models. Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1 for full model details). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. 
(In the main text we focus on the model with non-instantaneous pair formation, extra-pair contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix~S2 presents results on the effect of adding heterogeneity to other model variants.)
While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity. As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016,blanquart_transmission_virulence_2016}.

@@

Choosing the initial conditions for the simulations is challenging.
In many modeling studies,
researchers are primarily interested in equilibria (or other long-term
dynamical attractors such as limit cycles) and are exploring models
that have a single stable attractor, so initial conditions can be
ignored as long as we run models for long enough.  In eco-evolutionary
dynamics, however, the initial conditions do affect our
conclusions. We have no empirical information that would justify a
particular choice of the fraction infected and the mean
and variance of the distribution of SPVL at the point when the
pandemic strain of HIV-1 entered the human population; in any case,
the level of realism of our model does not support such a detailed
consideration of the early dynamics of HIV.  In most cases, we started
with an initial \Lspvl\ of 3.0, to match the value used by
Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
because we calibrated parameters based on the initial epidemic growth rate
(see ``Simulation runs'' below), we set $I(0)$ to $10^{-4}$
for most runs to ensure
that the exponential growth phase lasted long enough for reliable
estimation of the initial growth rate.

Appendix~S1 provides
further details on initial conditions, such
as the initial distribution of SPVL around the mean and the
distribution of initial infected density across single people
and different partnership types.

@@

We ran most of our models across a wide range of parameters, as described
in the next (\emph{Latin hypercube sampling}) 
section. In several cases, however, we inspected only a few
parameter sets, to qualitatively assess the sensitivity of the models to
initial conditions or to model structure. 
In particular, we tested model sensitivity 
to the initial prevalence $I(0)$ and initial mean SPVL
$\alpha(0)$, using baseline values of all parameters (Table~\ref{table:parmsTable}).
Using baseline parameter values,
we also ran all four basic model structure (\figurename~\ref{fig:diagrams})
with vital dynamics and with heterogeneity in sexual contact,
to assess the sensitivity of our results to these phenomena.

@@

One new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see Appendix S1). Since it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  we need to pick a reasonable range for it. Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.  Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5. Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$. Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is known about the uncoupled transmission rate.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3.0, standard deviation of \Lspvl\ of 0.2, and epidemic size of $10^{-4}$. After each run, the initial exponential growth rate is calculated. Then the parameters are scaled as described above so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
When calibrating, we run each model for only 500 years
(full simulations are run for 4000 years), which is always long
enough to capture the exponential growth phase of the model. 
We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations. 
(For the heterogeneous model, approximately
10\% of the samples failed due to numerical instability;
we discarded these runs.)

@@

Our chosen summary statistics (peak time, maximum mean \Lspvl, equilibrium mean \Lspvl, and minimum mean progression time) 
all vary considerably across models
(\figurename~\ref{fig:unidist}).
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. 
Some parameter
sets for these models lead to low equilibrium virulence ($2.3$-$3$ \Lspvl).
For these data sets, virulence may either increase from its initial value,
reaching an early peak ($\approx 200$ years) between 3 and 4 \Lspvl
and then declining to a lower equilibrium value, or in extreme cases
virulence may decline immediately, leading to a peak virulence (as we have
defined it) equal to the starting value of $\alpha(0)=3$ \Lspvl at $t=0$
(\figurename~\ref{fig:pairplot}).
At the opposite extreme, parameter sets that produce high equilibrium virulence (4.7 \Lspvl)
also produce late peaks ($> 200~\text{years}$) and
high peak virulence (5.6 \Lspvl).

@@

The random-mixing, pairform+epc,
and heterogeneous models all predict high population mean \Lspvl\ at the virulence peak
(median (95\% CI) = 4.7 (4.65-4.79), 
4.72 (4.37-4.96), 4.72  (4.09-5.03)). 
In contrast, the implicit model predicts a much lower peak \Lspvl\ value:
3.52 (3-4.02) years.
The random-mixing, pairform+epc, and heterogeneous models predict rapid
progression to AIDS at the virulence peak
(median/95\% CI = 6.1 (5.7-6.3), 6.02 (5.04-7.7), 6.03  (4.8-9.2)),
while the implicit model predicts minimum progression times about twice as long
(12.5 (9.6-15.6) years).
The corresponding differences in mean within-couple transmission
probability at the peak are even more extreme, about a fourfold difference:
0.249 (0.24-0.26), 0.252 (0.19-0.28), and 0.252 (0.15-0.28) per year for the 
random and pairform+epc models vs. 0.059 (0.02-0.13) per year
for the implicit model. (Appendix S2 
presents plots showing univariate summaries
of expected progression time to AIDS and transmission probability.)

@@

Finally, incorporating additional realism to the model, i.e.
combining heterogeneity with all four basic contact structures or
allowing for vital dynamics rather than assuming an SIS model,
leads to only small differences in the conclusions stated so far
(\figurename~S4). Relative to our baseline SIS assumption, the
effect of adding vital dynamics is to delay the virulence peak
slightly and increase both the peak and equilibrium virulence.
The changes are small, however: across all models, the
maximum increase in time until
the virulence peak is 40 years (for the instswitch+epc model),in 
the peak \Lspvl\ is 0.24 units (instswitch), and in the equilibrium
\Lspvl\ is 0.4 (pairform). The changes in the most realistic
model (pairform+epc) are considerably smaller: 
an increase of 2 years vs. a decrease of 13 years in the
time to the virulence peak for the
models with vital dynamics and heterogeneity, respectively;
an increase of 0.1 units vs. a decrease of 0.01 units in peak \Lspvl;
and an increase of 0.145 units vs. a decrease of 0.03 units
in equilibrium~\Lspvl. Thus, while we can never rule out the
possibility of some higher-order interaction among epidemiological
phenomena leading to significant changes in our conclusions,
we are reasonably confident that the results reported here are
robust to additional complexities.

@@

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{Appendix S1: Model Details}
\label{S1_Appendix}

\paragraph*{Appendix S2: Dynamics of Transmission and Virulence}
\label{S2_Appendix}

>>>>

In models that allow extra-pair contact,
the uncoupled and extra-couple transmission rates (i.e., the rates of
transmission among people outside of a stable partnership, or between
people inside of a stable partnership and people other than their
partner) are scaled by
multiplying the within-couple transmission rate $\beta$ by the contact
ratios $c_u/c_w$ and $c_e/c_w$ (see S1 Appendix).

@@

Lastly, we incorporated heterogeneity in sexual activity into the models. Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see S1 Appendix for full model details). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. 
(In the main text we focus on the model with non-instantaneous pair formation, extra-pair contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~D in S2 Appendix presents results on the effect of adding heterogeneity to other model variants.)
While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity. As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016,blanquart_transmission_virulence_2016}.

@@

Choosing the initial conditions for the simulations is challenging.
In many modeling studies,
researchers are primarily interested in equilibria (or other long-term
dynamical attractors such as limit cycles) and are exploring models
that have a single stable attractor, so initial conditions can be
ignored as long as we run models for long enough.  In eco-evolutionary
dynamics, however, the initial conditions do affect our
conclusions. We have no empirical information that would justify a
particular choice of the fraction infected and the mean
and variance of the distribution of SPVL at the point when the
pandemic strain of HIV-1 entered the human population; in any case,
the level of realism of our model does not support such a detailed
consideration of the early dynamics of HIV.  In most cases, we started
with an initial \Lspvl\ of 3.0, to match the value used by
Shirreff \etal \cite{shirreff_transmission_2011}. Shirreff \etal\ use an initial prevalence $I(0)=10^{-3}$;
because we calibrated parameters based on the initial epidemic growth rate
(see ``Simulation runs'' below), we set $I(0)$ to $10^{-4}$
for most runs to ensure
that the exponential growth phase lasted long enough for reliable
estimation of the initial growth rate.

S1 Appendix provides
further details on initial conditions, such
as the initial distribution of SPVL around the mean and the
distribution of initial infected density across single people
and different partnership types.

@@

We ran most of our models across a wide range of parameters, as described
in the next (\emph{Latin hypercube sampling}) 
section. In several cases, however, we inspected only a few
parameter sets, to qualitatively assess the sensitivity of the models to
initial conditions or to model structure. 
In particular, we tested model sensitivity 
to the initial prevalence, $I(0)$, and initial mean \Lspvl,
$\alpha(0)$, using baseline values of all parameters (Table~\ref{table:parmsTable}).
Using baseline parameter values,
we also ran all four basic model structure (\figurename~\ref{fig:diagrams})
with vital dynamics and with heterogeneity in sexual contact,
to assess the sensitivity of our results to these phenomena.

@@

One new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see S1 Appendix). Since it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  we need to pick a reasonable range for it. Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.  Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5. Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$. Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is known about the uncoupled transmission rate.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3.0, standard deviation of \Lspvl\ of 0.2, and epidemic size of $10^{-4}$. After each run, the initial exponential growth rate is calculated. Then the parameters are scaled as described above so that the initial exponential growth rate is scaled to 0.04 \peryear, a value that approximates the growth rates of Shirreff \etal's original models.
When calibrating, we run each model for only 500 years
(full simulations are run for 4000 years), which is always long
enough to capture the exponential growth phase of the model. 
We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations. 
(For the heterogeneous model, approximately
10\% of the samples failed due to numerical instability;
we discarded these runs.)

@@

Our chosen summary statistics (peak time, maximum mean \Lspvl, equilibrium mean \Lspvl, and minimum mean progression time) 
all vary considerably across models
(\figurename~\ref{fig:unidist}).
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. 
Some parameter
sets for these models lead to low equilibrium virulence ($2.3$-$3$ \Lspvl).
For these data sets, virulence may either increase from its initial value,
reaching an early peak ($\approx 200$ years) between 3 and 4 \Lspvl
and then declining to a lower equilibrium value, or in extreme cases
virulence may decline immediately, leading to a peak virulence (as we have
defined it) equal to the starting value of $\alpha(0)=3$ \Lspvl\ at $t=0$
(\figurename~\ref{fig:pairplot}).
At the opposite extreme, parameter sets that produce high equilibrium virulence (4.7 \Lspvl)
also produce late peaks ($> 200~\text{years}$) and
high peak virulence (5.6 \Lspvl).

@@

The random-mixing, pairform+epc,
and heterogeneous models all predict high population mean \Lspvl\ at the virulence peak
(median (95\% CI) = 4.7 (4.65-4.79), 
4.72 (4.37-4.96), 4.72  (4.09-5.03)). 
In contrast, the implicit model predicts a much lower peak \Lspvl\ value:
3.52 (3-4.02) years.
The random-mixing, pairform+epc, and heterogeneous models predict rapid
progression to AIDS at the virulence peak
(median/95\% CI = 6.1 (5.7-6.3), 6.02 (5.04-7.7), 6.03  (4.8-9.2)),
while the implicit model predicts minimum progression times about twice as long
(12.5 (9.6-15.6) years).
The corresponding differences in mean within-couple transmission
probability at the peak are even more extreme, about a fourfold difference:
0.249 (0.24-0.26), 0.252 (0.19-0.28), and 0.252 (0.15-0.28) per year for the 
random and pairform+epc models vs. 0.059 (0.02-0.13) per year
for the implicit model. (S2 Appendix presents plots showing univariate summaries
of expected progression time to AIDS and transmission probability.)

@@

Finally, incorporating additional realism to the model, i.e.
combining heterogeneity with all four basic contact structures or
allowing for vital dynamics rather than assuming an SIS model,
leads to only small differences in the conclusions stated so far
(\figurename~D). Relative to our baseline SIS assumption, the
effect of adding vital dynamics is to delay the virulence peak
slightly and increase both the peak and equilibrium virulence.
The changes are small, however: across all models, the
maximum increase in time until
the virulence peak is 40 years (for the instswitch+epc model),in 
the peak \Lspvl\ is 0.24 units (instswitch), and in the equilibrium
\Lspvl\ is 0.4 (pairform). The changes in the most realistic
model (pairform+epc) are considerably smaller: 
an increase of 2 years vs. a decrease of 13 years in the
time to the virulence peak for the
models with vital dynamics and heterogeneity, respectively;
an increase of 0.1 units vs. a decrease of 0.01 units in peak \Lspvl;
and an increase of 0.145 units vs. a decrease of 0.03 units
in equilibrium~\Lspvl. Thus, while we can never rule out the
possibility of some higher-order interaction among epidemiological
phenomena leading to significant changes in our conclusions,
we are reasonably confident that the results reported here are
robust to additional complexities.

@@

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Appendix: Model Details}
\label{S1_Appendix}

\paragraph*{S2 Appendix: Dynamics of Transmission and Virulence}
\label{S2_Appendix}
